# Dotaro Font

[![Project Status: WIP – Initial development is in progress, but there has not yet been a stable, usable release suitable for the public.](https://www.repostatus.org/badges/latest/wip.svg)](https://www.repostatus.org/#wip)

**Warning:** this project is a Work-in-Progress.

```{r hidden, echo = FALSE}
knitr::opts_chunk$set(fig.path = "man/figures/README-", fig.cap = '', fig.height = 1.1)

```

### Table of Contents

* [Overview](#overview)
* [Downloadable font files](#downloads)
* [Examples](#examples)
* [Unicode block coverage](#blocks)
* [License](#license)
* [Build the fonts from scratch](#build)
* [Related links](#related)

## <a name="overview">Overview</a>

The main goal is to create a couple of fonts with bespoke glyphs to use in my [Dotaro Deck](https://github.com/trevorld/dotaro.deck) of playing cards:

### Dotaro Ranks

* A narrow monospace font to provide *rank* symbols in playing card [corner indices](https://www.wopc.co.uk/playing-cards/corner-indices).
* Should look good when rendered with a border and a colored fill with R's [`fillStrokeGrob()`](https://stat.ethz.ch/R-manual/R-devel/library/grid/html/grid.stroke.html) at a target size of 1/4&Prime; &times; 3/8&Prime; (corresponds to ~46pt).
* Mainly [slab serif](https://en.wikipedia.org/wiki/Slab_serif).

### Dotaro Suits

* A square monospace font to provide *suit* symbols in playing card corner indices and pips.
* Should look good when rendered with a border and a colored fill with R's [`fillStrokeGrob()`](https://stat.ethz.ch/R-manual/R-devel/library/grid/html/grid.stroke.html) at a target size of 1/4&Prime; &times; 1/4&Prime; (corresponds to ~23pt).

## <a name="downloads">Downloadable Font Files</a>

* [dotaro-ranks.ttf](https://trevorldavis.com/share/fonts/dotaro-ranks.ttf)
* [dotaro-suits.ttf](https://trevorldavis.com/share/fonts/dotaro-suits.ttf)
* [dotaro-ranks-code-charts.pdf](https://trevorldavis.com/share/fonts/dotaro-ranks-code-charts.pdf)
* [dotaro-suits-code-charts.pdf](https://trevorldavis.com/share/fonts/dotaro-suits-code-charts.pdf)

## <a name="examples">Examples</a>

**Note:** Although *Dotaro Ranks* and *Dotaro Suits* are **monochrome** fonts in the examples below I use R's [`fillStrokeGrob()`](https://stat.ethz.ch/R-manual/R-devel/library/grid/html/grid.stroke.html) to rendered colored versions of the glyphs with outlines.

### Traditional Anglo-American pattern ranks

```{r numerals, echo=FALSE, fig.alt = "The digits in Dotaro Ranks shown as card indices"}
df <- dotaro.font:::df_index(as.character(0:9))
dotaro.font:::grid.index(df, bleed = "white")
```
```{r face_letters, echo=FALSE, fig.alt = "Some capital letters in Dotaro Ranks shown as card indices"}
df <- dotaro.font:::df_index(c("\uf590", "J", "C", "Q", "K", "A"))
dotaro.font:::grid.index(df, bleed = "white")
```

### Chess ranks

```{r face_chess, echo=FALSE, fig.alt = "Some chess glyphs in Dotaro Ranks shown as card indices"}
df <- dotaro.font:::df_index(c("\u265f", "\u265e", "\u265d", "\u265c", "\u265b", "\u265a"))
dotaro.font:::grid.index(df, bleed = "white")
```
```{r chess_meeples, echo=FALSE, fig.alt = "Chess glyphs combined with meeple glyphs"}
library("grid")

# Unicode code points (Dotaro Square font)
MEEPLE_BLACK <- "\U000FC431"
PAWN_BLACK <- "\u265f"
KNIGHT_BLACK <- "\u265e"
BISHOP_BLACK <- "\u265d"
ROOK_BLACK <- "\u265c"
QUEEN_BLACK <- "\u265b"
KING_BLACK <- "\u265a"

# Font sizes in points
meeple_size <- 60
chess_size <- 26

# Layout dimensions (inches)
n_cols <- 6
col_spacing <- 0.85
total_width <- n_cols * col_spacing + 0.3
total_height <- 1.8

# x center of each column
xs <- unit(0.15 + col_spacing * (seq_len(n_cols) - 0.5), "in")

# Meeple y-center
y_meeple <- unit(0.7, "in")

# Hat y-center: just above the meeple's head
# Meeple em-box half-height = meeple_size / 2 pts; chess em-box half = chess_size / 2 pts
# Overlap slightly so the hat sits on the head rather than floating above
hat_offset_in <- (meeple_size / 2 + chess_size * 0.35) / 72 - 0.15
y_hat <- y_meeple + unit(hat_offset_in, "in")

# Knight y-center: just below the meeple (riding position)
knight_offset_in <- (meeple_size / 2 + chess_size * 0.35) / 72
y_knight <- y_meeple - unit(knight_offset_in, "in")

make_glyph <- function(glyph, x, y, size, col = "black", fill = "black") {
    tg <- textGrob(
        label = glyph,
        x = x,
        y = y,
        gp = gpar(fontsize = size, fontfamily = "Dotaro Square")
    )
    fillStrokeGrob(tg, gp = gpar(col = col, fill = fill, lwd = 1.5))
}

vp <- viewport(
    width = unit(total_width, "in"),
    height = unit(total_height, "in"),
    x = unit(0.4, "npc"),
    y = unit(0.5, "npc")
)
pushViewport(vp)
hats <- c(PAWN_BLACK, BISHOP_BLACK, QUEEN_BLACK, KING_BLACK)

# Meeples 1-4: chess piece worn as hat on top
for (i in seq_len(4)) {
	if (i < 2L) {
		x <- xs[1]
	} else if (i < 3L) {
		x <- xs[3]
	} else {
		x <- xs[i + 2L]
	}
    grid.draw(make_glyph(MEEPLE_BLACK, x, y_meeple, meeple_size, fill = "#555555"))
    grid.draw(make_glyph(hats[i], x, y_hat, chess_size, fill = "lightgrey"))
}

# Meeple 5: knight below, as if riding
grid.draw(make_glyph(MEEPLE_BLACK, xs[2], y_hat, 2.1 * chess_size, fill = "#555555"))
grid.draw(make_glyph(KNIGHT_BLACK, xs[2], y_meeple, 0.9 * meeple_size, fill = "lightgrey"))

# Meeple 6: rook below, as if standing on top
grid.draw(make_glyph(MEEPLE_BLACK, xs[4], y_hat, 1.5 * chess_size, fill = "#555555"))
grid.draw(make_glyph(ROOK_BLACK, xs[4], y_meeple, 1.0 * meeple_size, fill = "lightgrey"))
popViewport()
```

## <a name="blocks">Unicode block coverage</a>

Current [Unicode block](https://en.wikipedia.org/wiki/Unicode_block) coverage (excluding control codes):

```{r blocks, echo=FALSE, eval=requireNamespace("Unicode", quietly = TRUE), results='asis'}
reticulate::use_python(Sys.which("python3"))
library("Unicode")
# assumes fonts in `share` directory
stopifnot(file.exists("share/dotaro-suits.ttf"),
          file.exists("share/dotaro-ranks.ttf"))
points_suits <- dotaro.font:::get_unicode_code_points("share/dotaro-suits.ttf") |> as.u_char()
points_ranks <- dotaro.font:::get_unicode_code_points("share/dotaro-ranks.ttf") |> as.u_char()
blocks <- u_blocks()
n_per_block <- lapply(blocks, n_of_u_chars)
n_per_block[["Basic Latin"]] <- 128L - 33L # Don't count control codes
n_per_block[["Latin-1 Supplement"]] <- 128L - 34L # Don't count control codes

n_per_block_suits <- lapply(blocks, \(x) sum(points_suits %in% as.u_char(x)))
n_per_block_ranks <- lapply(blocks, \(x) sum(points_ranks %in% as.u_char(x)))
positive_suits <- Filter(as.logical, n_per_block_suits) |> names()
positive_ranks <- Filter(as.logical, n_per_block_ranks) |> names()
shared_blocks <- union(positive_suits, positive_ranks)

n_per_block <- n_per_block[shared_blocks]
n_per_block_suits <- n_per_block_suits[shared_blocks]
n_per_block_ranks <- n_per_block_ranks[shared_blocks]
df <- data.frame(Block = names(n_per_block),
                 `Dotaro Ranks` = as.integer(n_per_block_ranks),
                 `Dotaro Suits` = as.integer(n_per_block_suits),
                 `Total in Block` = as.integer(n_per_block), check.names = FALSE)
knitr::kable(df)
```

## <a name="license">License</a>

This R package is licensed under the MIT license.

The Dotaro Font files generated by this R package are licensed under the [SIL Open Font License, Version 1.1](https://openfontlicense.org/).

## <a name="build">Build the fonts from scratch</a>

Install the R package with

```r
remotes::install_github("trevorld/dotaro.font")
```

Also install [FontForge](https://fontforge.org/en-US/) and its python bindings e.g. `apt install fontforge python3-fontforge` on Debian-based Linux distributions.

Run the following R code

```r
library("dotaro.font")
reticulate::use_python(Sys.which("python3"))

generate_sfd("ranks", "dotaro-ranks.sfd")
generate_font("dotaro-ranks.sfd", "dotaro-ranks.ttf")

generate_sfd("suits", "dotaro-suits.sfd")
generate_font("dotaro-suits.sfd", "dotaro-suits.ttf")
```

**Warning:** I've noticed it may not update a previously generated font unless run in a clean R session.

## <a name="related">Related links</a>

### FontForge:

* [FontForge Site](https://fontforge.org/en-US/)

  + [Python Scripting](https://fontforge.org/docs/scripting/python.html)
  + [Technical References](https://fontforge.org/docs/techref.html)

* [Design With FontForge](http://designwithfontforge.com/en-US/index.html)

  + [The Final Output, Generating Font Files](http://designwithfontforge.com/en-US/The_Final_Output_Generating_Font_Files.html)
  + [Importing Glyphs from Other Programs](http://designwithfontforge.com/en-US/Importing_Glyphs_from_Other_Programs.html)

* [Creating Fonts with Inkscape and FontForge](https://www.reddit.com/r/neography/comments/818364/creating_fonts_with_inkscape_and_fontforge_table/)

  + [FontForge Scripting](https://www.reddit.com/r/neography/comments/83ovk7/creating_fonts_with_inkscape_and_fontforge_part10/)

### Fonts and glyphs

#### Chess fonts and glyphs

* [chessvariants.com: Chess Word Macros & Fonts](https://www.chessvariants.com/d.font/fonts.html)
* [chessvariants.com: Unicode Fonts with Chess Piece Images](https://www.chessvariants.com/d.font/unicode.html)
* [enpassant.dk: Chess Fonts](https://www.enpassant.dk/chess/fonteng.htm)
* [lichess.org: svg images](https://github.com/lichess-org/lila/tree/master/public/piece)
* [Luc Devroye: Chess fonts](https://luc.devroye.org/chess.html)

#### Playing card fonts and glyphs

Collections of links:

* [Luc Devroye: Playing card fonts](https://luc.devroye.org/cards.html)
* [pagat.com links: Bitmaps, fonts, programming tools](https://www.pagat.com/links/images.html)

Select playing card fonts:

* [Card Characters](https://www.haroldsfonts.com/font/card-characters)
* [Cards Font](https://www.fontspace.com/cards-font-f3690)
* [Cards TrueType Font](https://www.pagat.com/com/cardsttf.html)
* [Fortune Letters and Widgets](https://www.fontmonkey.com/archive.php?font=fortune) (i.e. the [Decktet](https://www.decktet.com/) fonts)
* [Hoyle Playing Cards](https://www.fontspace.com/hoyle-playing-cards-font-f3686)
* [JQKAs Wild](https://www.fontspace.com/jqkas-wild-font-f96423)
* [Numbers: Deuce](https://www.typography.com/fonts/numbers/)

#### Other Slab Serif fonts

* [Amazing Slab](https://www.dafont.com/amazing-slab.font)
* [Arvo](https://fonts.google.com/specimen/Arvo)
* [Bondie Condensed Slab Serif](https://craftsupply.co/product/bondie-condensed-slab-serif-font/)
* [Cabo Slad](https://www.dafont.com/cabo-slab.font)
* [Jerome Condensed Slab Serif](https://www.dafont.com/jerome.font)
* [Luna](https://creativemarket.com/orcacreative/866559-Luna)
* [Memphis](https://en.wikipedia.org/wiki/Memphis_(typeface))
* [Neutraface Slab](https://housefonts.com/hi/neutraface_slab)
* [Rockwell](https://en.wikipedia.org/wiki/Rockwell_(typeface)) (i.e. [Geometric Slabserif 712](https://www.fontspring.com/fonts/paratype/geometric-slabserif-712?srsltid=AfmBOoof4JDryBjju-n4XtArWi5hKfHLe0PSqIyud8oR8rJApCCd3gJC))
* [Servat](https://fontlot.com/servat-font/)
* [Stymie](https://www.youworkforthem.com/font/T1050/stymie/typetester/)
* [Winchester Condensed](https://www.creativefabrica.com/product/winchester-condensed/)
* [Zilla Slab](https://fonts.google.com/specimen/Zilla+Slab/tester) ([blog post](https://blog.mozilla.org/opendesign/zilla-slab-common-language-shared-font/))

#### Square fonts

* [Alpine Fonts: Chess and Other Games](https://partae.com/fonts/index.php/general-info)
* [Game Bit Mono](https://github.com/trevorld/game-bit-font/)
* [Kreative Square](https://www.kreativekorp.com/software/fonts/ksquare/)

#### Other fonts

* [Catrinity](https://catrinity-font.de/)
* [Fairfax HD](https://www.kreativekorp.com/software/fonts/fairfaxhd/)
* [Nishiki-teki](https://umihotaru.work/)
* [Quivira](http://quivira-font.com/)
